The Leaky Semicolon

### 2021-10-26 paper/mca1.tex
removing todos

In this section, we develop a model of concurrent computation by adding
% \emph{parallel composition} and 
\emph{reads-from} to \reffig{fig:seq}.  To model coherence
and synchronization, we add \emph{delay} to the rule for sequential
composition.  For  \mca{} architectures, it is sufficient to encode delay in
the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
for relaxed access on \armeight{}, but requires extra synchronization for
acquiring reads.  (\emph{Lowering} is the translation of language-level
operators to machine instructions.  A lowering is \emph{optimal} if it
provides the most efficient execution possible.)

@@

Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  For further
discussion of item \eqref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
see \cite{appendix}.
%see \textsection\ref{sec:delay}.

>>>>

In this section, we develop a model of concurrent computation by adding
% \emph{parallel composition} and 
\emph{reads-from} to \reffig{fig:seq}.  To model coherence and
synchronization, we add \emph{delay} to the rule for sequential composition.
For \mca{} architectures, it is sufficient to encode delay in the pomset
order.  The resulting model, \PwTmca{1}, supports optimal lowering for
relaxed access on \armeight{}, but requires extra synchronization for
acquiring reads.\footnote{\citet{DBLP:journals/pacmpl/JagadeesanJR20}
  erroneously elide the required synchronization on acquiring reads.}
(\emph{Lowering} is the translation of language-level operators to machine
instructions.  A lowering is \emph{optimal} if it provides the most efficient
execution possible.)

@@

Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  We discuss
\ref{lem:if}\ref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
further in \textsection\ref{sec:outro}.

### 2021-10-26 paper/oopsla.tex
Updates
removing todos

Contrary to the claim, redundant read elimination fails for \jjr{}.
We discuss redundant read elimination in \textsection\ref{sec:semreg}.
Consider JMM Causality Test Case 2, which we describe there.
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \PAR
  \PW{x}{y}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{\DW{y}{1}}{right=of a2}
      \event{b1}{\DR{y}{1}}{right=3em of a3}
      \event{b2}{\DW{x}{1}}{right=of b1}
      \rf{a3}{b1}
      \po{b1}{b2}
      \rf[out=169,in=11]{b2}{a2}
      \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}
Under the semantics of \jjr{}, we have
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\bigmid\DW{y}{1}}{right=of a2}
    \end{tikzinline}}
\end{gather*}
The precondition of $\DWP{y}{1}$ is \emph{not} a tautology, and therefore
redundant read elimination fails.
(It is a tautology in
\begin{math}
  \PR{x}{r}\SEMI
  \LET{s}{r}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
\end{math}.)
\jjr{\textsection3.1} incorrectly stated that the precondition of
$\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.

\myparagraph{Parallel Composition}

In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
of events.  Here we have forbidden coalescing.  This difference appears to be
arbitrary.  In \jjr{}, however, there is a mistake in the handling of
termination actions.  The predicates should be joined using $\land$, not
$\lor$.

@@

\myparagraph{Augmentation of Preconditions}
\jjr{} allows augmentation of preconditions.
As discussed in \textsection\ref{sec:delay}, this causes associativity to fail
for $\rdelay$, at least when attempting to validate \reflem{lem:if}\eqref{lem:ifelse:if:if1}--\eqref{lem:ifelse:if:if2}
Thus, we use \emph{weakest} preconditions, rather than general preconditions.
As a result, we fail to validate the following
refinement:
\begin{math}
  \aPSS_1
  \not\supseteq
  \xIFTHEN{\aForm}{\aPSS_1}{}.
\end{math}

>>>>

Contrary to the claim, redundant read elimination fails for \jjr{}.
We discuss redundant read elimination in \textsection\ref{sec:semreg}.
Consider JMM Causality Test Case 2, which we describe there.
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \PAR
  \PW{x}{y}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{\DW{y}{1}}{right=of a2}
      \event{b1}{\DR{y}{1}}{right=3em of a3}
      \event{b2}{\DW{x}{1}}{right=of b1}
      \rf{a3}{b1}
      \po{b1}{b2}
      \rf[out=-169,in=-11]{b2}{a2}
      \rf[out=-169,in=-11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}
Under the semantics of \jjr{}, we have
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\bigmid\DW{y}{1}}{right=of a2}
    \end{tikzinline}}
\end{gather*}
The precondition of $\DWP{y}{1}$ is \emph{not} a tautology, and therefore
redundant read elimination fails.
(It is a tautology in
\begin{math}
  \PR{x}{r}\SEMI
  \LET{s}{r}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
\end{math}.)
\jjr{\textsection3.1} incorrectly stated that the precondition of
$\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.

\myparagraph{Termination Conditions and Parallel Composition}

In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
of events.  Here we have forbidden coalescing.  This difference appears to be
arbitrary.  In \jjr{}, however, there is a mistake in the handling of
termination actions.  The predicates should be joined using $\land$, not
$\lor$.  Here we have used termination conditions rather than termination
actions so that termination is handled separately.

@@

\myparagraph{Augmentation of Preconditions}
\jjr{} allows arbitrary augmentation of preconditions.  Here we are more
conservative, only allowing augmentation of preconditions in the semantics of
primitive actions, as in \textsection\ref{sec:semca}.
As discussed in \textsection\ref{sec:delay}, allowing arbitrary augmentation
causes associativity to fail when encoding $\rdelay$ logically. 
% Thus, we use \emph{weakest} preconditions, rather than general preconditions.
% As a result, we fail to validate the following
% refinement:
% \begin{math}
%   \aPSS_1
%   \not\supseteq
%   \xIFTHEN{\aForm}{\aPSS_1}{}.
% \end{math}

### 2021-10-26 paper/outro.tex
removing todos

% In a similar vein,
\PwT{} does not validate read introduction, whereas speculative operational
semantics do.  Recent work shows a tension between read introduction and
compositional reasoning for temporal safety properties (see
\textsection\ref{sec:related}).  Nonetheless, read introduction is
ubiquitous in some compilers.  It would be interesting to know if there is a
performance penalty for banning read introduction.

>>>>

% In a similar vein,
% \PwT{} does not validate read introduction, whereas speculative operational
% semantics do.  Recent work shows a tension between read introduction and
% compositional reasoning for temporal safety properties (see
% \textsection\ref{sec:related}).  Nonetheless, read introduction is
% ubiquitous in some compilers.  It would be interesting to know if there is a
% performance penalty for banning read introduction.

### 2021-10-26 paper/related.tex
removing todos

\todo{Text about PwP.}

\todo{By the way, I note that PwP doesn't use these termination conditions,
  and there's no explanation in the paper of why this part of the model has
  changed, even in the appendix -- could you add it?}

% We provide a detailed comparison with
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.

@@

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
%
Many of the most prominent models are operational, whole-program models based
on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
  DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
% We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.

\todo{We agree that a new memory model needs to be positioned against existing
models.  The usual result here is a compilation correctness to hardware
memory models.  For PwT-MCA, we address this by showing compilation result
for Armv8 model (§5).
%
Comparing software models, however, is unsatisfying: they are all
incomparable, i.e., there are examples which are allowed by one/disallowed by
the other and vice versa.
%
Morally, our model sits between the strong models (exemplified by RC11
[Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
semantics [Kang-al:POPL17]).  As we argue in §1:
%
- The strong models require too much synchronization.
%
- The speculative models allow thin air behaviors.
%
Looking at the details, however, PwT-MCA is incomparable to both RC11 and
promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
PwT-MCA has a weaker notion of coherence than the promising semantics.
%
Some differences reflect an attempt to fix a bug.  For example, Weakestmo
[Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
behaviors of the promising semantics.
%
Other differences reflect a different balance between allowed optimizations
and reasoning principles.  There are fundamental conflicts, for example:
%
- between Common Subexpression Elimination (CSE) and read-read coherence
  (§D.1)
%
- between if-introduction (aka, case analysis, if-closure) and java-style
  final-field semantics.  (If-introduction requires that address dependencies
  and control dependencies are the same.  Final-fields require that they be
  different.)
}
\todo{Text about promising.}

>>>>

There are many differences between \PwP{}
\cite{DBLP:journals/pacmpl/JagadeesanJR20} and \PwT{}.  Two of these are
motivated by requirements unique to \PwT{}: First, \PwP{} uses substitution
to define read actions; \PwT{} Skolemizes in order to satisfy disjunction
closure for predicate transformers (\textsection\ref{sec:ex:assoc}).  Second,
\PwP{} requires \emph{consistency} and \emph{causal strengthening}; \PwT{}
drops these in order to satisfy associativity (\reflem{lem:monoid}).  Other
differences are stylistic: For example, we replaced the termination
\emph{actions} of \PwP{} by termination \emph{conditions}---our formulation
fixes an error in \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20}'s
definition of parallel composition.  We also fix an error in their treatment
of redundant read elimination.

@@

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
%
Many of the most prominent models are operational models based
on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
  DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
Morally, \PwT{} fits between the strong models and the speculative ones.
% We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.

% \todo{We agree that a new memory model needs to be positioned against existing
% models.  The usual result here is a compilation correctness to hardware
% memory models.  For PwT-MCA, we address this by showing compilation result
% for Armv8 model (§5).
% %
% Comparing software models, however, is unsatisfying: they are all
% incomparable, i.e., there are examples which are allowed by one/disallowed by
% the other and vice versa.
% %
% Morally, our model sits between the strong models (exemplified by RC11
% [Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
% semantics [Kang-al:POPL17]).  As we argue in §1:
% %
% - The strong models require too much synchronization.
% %
% - The speculative models allow thin air behaviors.
% %
% Looking at the details, however, PwT-MCA is incomparable to both RC11 and
% promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
% PwT-MCA has a weaker notion of coherence than the promising semantics.
% %
% Some differences reflect an attempt to fix a bug.  For example, Weakestmo
% [Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
% behaviors of the promising semantics.
% %
% Other differences reflect a different balance between allowed optimizations
% and reasoning principles.  There are fundamental conflicts, for example:
% %
% - between Common Subexpression Elimination (CSE) and read-read coherence
%   (§D.1)
% %
% - between if-introduction (aka, case analysis, if-closure) and java-style
%   final-field semantics.  (If-introduction requires that address dependencies
%   and control dependencies are the same.  Final-fields require that they be
%   different.)
% }
\citet{DBLP:journals/pacmpl/JagadeesanJR20} argue that the speculative models
allow too many executions, resulting in a failure of temporal reasoning and
potentially jeopardizing type safety and other security properties.  In a
similar vein, \citet{promising-ldrf} argue that local \drf{} guarantees are
violated when read-introduction is followed by if-introduction, branching on
the read just introduced.  These optimizations are validated by the
speculative models---\citeauthor{promising-ldrf} manage to avoid the problem
by adopting a sub-optimal lowering for \RMW{}s.  \PwT{} does not suffer from
this problem, since \PwT{} does not validate read-introduction.  Nonetheless,
read introduction is ubiquitous in some compilers
\cite{DBLP:conf/pldi/LeeKSHDMRL17}.  There appears to be a genuine tension
between temporal reasoning, as supported by \PwT{}, and read-introduction, as
supported by the speculative models.
